Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-05-30 paper-amia-2018/manuscript.tex
update manuscript
update figure 1 and results for RNNs

\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}). Our GRU model includes one hidden layer, output layer and input layer. We reset our model state after feeding each input sequence where input was given as one-hot encoding of word vector. Since LSTM and GRU use one-hot vector as input, results for these models are reported only when lexical and punctuation features are used. We experimentally determined that the best performance is achieved when the number of hidden units is 25, batch size is 1 and Adam\cite{kingma2014adam} is used for optimization.

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.762 & \textbf{0.673} & 0.715 & \textbf{0.989} & 0.993 & 0.991  & 0.981 & 0.982 & 0.982 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & \textbf{0.989} & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.800} & 0.646 & 0.714  & 0.993 & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & 0.983 & 0.984 \\ \hline
 GRU & \textbf{0.769} & \textbf{0.715} & \textbf{0.741} & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.986} & \textbf{0.985} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and weighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.673 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\% and 36.26\% higher precision, recall and F1-measure, respectively, in ``same segment'' identification compared to ``new segment'' detection. When lexical features are used in combination with punctuation mark, GRU demonstrates the best performance in all cases except precision for ``new segment'' identification where LSTM shows 4.03\% higher precision than GRU. In contrast, GRU shows 10.68\% higher recall and 3.78\%  higher F1-measure than LSTM for ``new segment'' identification, 0.1\% higher precision than LSTM for ``same segment'' identification and 0.2\% higher recall and 0.2\%  higher F1-measure than LSTM for overall classification. \\

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows the second highest AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.833 AUCs when only lexical features are used; and 0.966 and 0.922 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.4\% lower AUC than LSTM.

>>>>

\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\label{eq:resetgru}
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\label{eq:candidategru}
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product. The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors. The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}). Our GRU model includes one hidden layer, output layer and input layer. We reset our model state after feeding each input sequence where input was given as one-hot encoding of word vector. Since LSTM and GRU use one-hot vector as input, results for these models are reported only when lexical and punctuation features are used. We experimentally determined that the best performance is achieved when the number of hidden units is 25, batch size is 1 and Adam\cite{kingma2014adam} is used for optimization. Source codes of this work will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication.

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.762 & \textbf{0.673} & 0.715 & \textbf{0.989} & 0.993 & 0.991  & 0.981 & 0.982 & 0.982 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & \textbf{0.989} & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.810} & 0.720 & \textbf{0.762}  & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
 GRU & 0.783 & \textbf{0.739} & 0.760 & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and weighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.673 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\% and 36.26\% higher precision, recall and F1-measure, respectively, in ``same segment'' identification compared to ``new segment'' detection. When lexical features are used in combination with punctuation mark, LSTM demonstrates the best performance in terms of precision and F1-measure for ``new segment'' identification while GRU shows 2.64\% higher recall than LSTM. GRU and LSTM show exactly same performance in terms of all performance metrics for ``same segement'' and overall classifications. \\

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows the second highest AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.833 AUCs when only lexical features are used; and 0.966 and 0.922 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.2\% lower AUC than LSTM.
